---
title: "Lab 2: Spatial Analysis and Visualization"
subtitle: "Healthcare Access and Equity in Pennsylvania"
author: "Mackenna Amole"
date: today
format: 
  html:
    code-fold: false
    toc: true
    toc-location: left
    theme: cosmo
    embed-resources: true
execute:
  warning: false
  message: false
---

## Assignment Overview

**Learning Objectives:**
- Apply spatial operations to answer policy-relevant research questions
- Integrate census demographic data with spatial analysis
- Create publication-quality visualizations and maps
- Work with spatial data from multiple sources
- Communicate findings effectively for policy audiences

---

## Part 1: Healthcare Access for Vulnerable Populations

### Research Question

**Which Pennsylvania counties have the highest proportion of vulnerable populations (elderly + low-income) living far from hospitals?**

Your analysis should identify counties that should be priorities for healthcare investment and policy intervention.

#### Step 1: Data Collection (5 points)

Load the required spatial data:
- Pennsylvania county boundaries
- Pennsylvania hospitals (from lecture data)
- Pennsylvania census tracts

**Your Task:**
```{r}
#| message: false
#| warning: false
#| results: 'hide'

# Load required packages
library(sf)
library(tidyverse)
library(tigris)
library(tidycensus)
library(scales)
library(patchwork)
library(here)

# Load spatial data
pa_counties <- st_read(here("labs/lab_2/CPLN-5920-SP26-main/lectures/week_04/data/Pennsylvania_County_Boundaries.shp"))
#pa_counties <- st_read(here("CPLN-5920-SP26-main/CPLN-5920-SP26-main/lectures/week_04/data/Pennsylvania_County_Boundaries.shp"))
#districts <- st_read(here("CPLN-5920-SP26-main/CPLN-5920-SP26-main/lectures/week_04/data/districts.geojson"))
districts <- st_read(here("labs/lab_2/CPLN-5920-SP26-main/lectures/week_04/data/districts.geojson"))
#hospitals <- st_read(here("CPLN-5920-SP26-main/CPLN-5920-SP26-main/lectures/week_04/data/hospitals.geojson"))
hospitals <- st_read(here("labs/lab_2/CPLN-5920-SP26-main/lectures/week_04/data/hospitals.geojson"))
census_tracts <- tracts(state = "PA", cb = TRUE)
metro_areas <- core_based_statistical_areas(cb = TRUE)



# Check that all data loaded correctly
glimpse(pa_counties)
glimpse(districts)
glimpse(hospitals)


# Check coordinates
st_crs(pa_counties)
st_crs(districts)
st_crs(hospitals)
st_crs(metro_areas)
st_crs(census_tracts)

```

**Questions**
- How many hospitals are in your dataset? 
  223
  
- How many census tracts? 
  3,445
  
- What coordinate reference system is each dataset in? 
  WGS84, except for metro_areas and census_tracts which are in NAD83

---

#### Step 2: Get Demographic Data 

Use `tidycensus` to download tract-level demographic data for Pennsylvania.

**Required variables:**
- Total population
- Median household income
- Population 65 years and over (you may need to sum multiple age categories)

**Your Task:**
```{r}
#| message: false
#| warning: false
#| results: 'hide'

# Get demographic data from ACS

# Get demographic data from ACS (tract-level, all of PA)
demo_vars <- c(
  total_pop = "B01003_001",
  med_income = "B19013_001",
  male_65_66 = "B01001_020",
  male_67_69 = "B01001_021",
  male_70_74 = "B01001_022",
  male_75_79 = "B01001_023",
  male_80_84 = "B01001_024",
  male_85_plus = "B01001_025",
  female_65_66 = "B01001_044",
  female_67_69 = "B01001_045",
  female_70_74 = "B01001_046",
  female_75_79 = "B01001_047",
  female_80_84 = "B01001_048",
  female_85_plus = "B01001_049"
)
tract_demographics <- get_acs(
  geography = "tract",
  state = "PA",
  variables = demo_vars,
  year = 2022,
  survey = "acs5",
  geometry = TRUE,
  output = "wide"
)
tract_demographics <- tract_demographics %>%
  mutate(
    pop_65_plus =
      male_65_66E + male_67_69E + male_70_74E +
      male_75_79E + male_80_84E + male_85_plusE +
      female_65_66E + female_67_69E + female_70_74E +
      female_75_79E + female_80_84E + female_85_plusE
  )
#tract demographics already has geometry

#tracts with missing income
sum(is.na(tract_demographics$med_incomeE))

#median income PA tracts
median(tract_demographics$med_incomeE, na.rm = TRUE)

```

**Questions**
- What year of ACS data are you using? 
  2022
  
- How many tracts have missing income data? 
  63
  
- What is the median income across all PA census tracts? 
  $70,188

---

#### Step 3: Define Vulnerable Populations 

Identify census tracts with vulnerable populations based on TWO criteria:
1. Low median household income 
2. Significant elderly population

**Your Task:**
```{r}
#| message: false
#| warning: false
#| results: 'hide'

#identifying vulnerable: $45k b/c less than median but more than poverty level in PA, 25% b/c PA average is about 20% so this represents a higher concentration
tract_demographics <- tract_demographics %>%
  mutate(
    pct_65_plus = pop_65_plus / total_popE,
    vulnerable = if_else(
      med_incomeE < 45000 & pct_65_plus > 0.25,
      TRUE, FALSE
    )
  )


# Filter for vulnerable tracts based on your criteria

vulnerable_tracts <- tract_demographics %>%
  filter(vulnerable == TRUE)


```

**Questions**
- What income threshold did you choose and why? 
  I chose 45,000 as the median income threshold. The PA tract average was just slightly over 70,000 and the PA poverty level is approx. 20,000 for 2-person household/approx. 25,000 for 3-person household; 45,000 falls in the middle of this range, highlighting people who are low-income and below state median but not necessarily living below the poverty level.
  
- What elderly population threshold did you choose and why? 
  I chose 25% age 65+. The PA average population over 65 is about 20% (according to a quick google search for the state scale), so I raised this slightly to identify places with even higher than average concentrations.

- How many tracts meet your vulnerability criteria? 
  78
  
- What percentage of PA census tracts are considered vulnerable by your definition? 
  approx. 2.26%

---

#### Step 4: Calculate Distance to Hospitals 

For each vulnerable tract, calculate the distance to the nearest hospital.

**Your Task:**
```{r}
#| message: false
#| warning: false
#| results: 'hide'

# Transform to appropriate projected CRS
vulnerable_proj <- st_transform(vulnerable_tracts, 2272)
hospitals_proj <- st_transform(hospitals, 2272)

#tract centroids
tract_centroids <- st_centroid(vulnerable_proj)

#distance matrix to all hospitals
distance_matrix <- st_distance(tract_centroids, hospitals_proj)

#min dist to nearest hospital in ft
nearest_distance_ft <- apply(distance_matrix, 1, min)

#ft to miles
vulnerable_proj$nearest_hospital_miles <- as.numeric(nearest_distance_ft) / 5280

#avg dist
mean_distance <- mean(vulnerable_proj$nearest_hospital_miles, na.rm = TRUE)
mean_distance

#max dist
max_distance <- max(vulnerable_proj$nearest_hospital_miles, na.rm = TRUE)
max_distance

#number of tracts > 15 miles from hospital
num_far <- sum(vulnerable_proj$nearest_hospital_miles > 15)
num_far

```

**Requirements:**
- Use an appropriate projected coordinate system for Pennsylvania
- Calculate distances in miles
- Explain why you chose your projection

**Questions**
- What is the average distance to the nearest hospital for vulnerable tracts? 
  approx. 3.88 miles
  
- What is the maximum distance? 
  18.13 miles
  
- How many vulnerable tracts are more than 15 miles from the nearest hospital? 
  3
  
- Which projection did you choose and why? 
  I transformed the data to Pennsylvania State Plane South (EPSG:2272). This is a projected coordinate system designed specifically for Pennsylvania (the southern half of the state, so chose south just because Philadelphia and Pittsburgh are both in the southern half).

---

#### Step 5: Identify Underserved Areas 

Define "underserved" as vulnerable tracts that are more than 15 miles from the nearest hospital.

**Your Task:**
```{r}
#| message: false
#| warning: false
#| results: 'hide'

# Create underserved variable
vulnerable_proj <- vulnerable_proj %>%
  mutate(underserved = nearest_hospital_miles > 15)

#number underserved tracts
num_underserved <- sum(vulnerable_proj$underserved)
num_underserved

#% vulnerable tracts that are underserved
pct_underserved <- (3/78 *100)
pct_underserved
```

**Questions to answer:** 
- How many tracts are underserved? 
  3
  
- What percentage of vulnerable tracts are underserved? 
  3.85%
  
- Does this surprise you? Why or why not? 
  This number feels low to me, especially with 78 vulnerable tracts identified. However, it does seem like a good thing to know that even though vulnerable tracts exist across the state, there are not nearly as many which are a far distance from a hospital.

---

#### Step 6: Aggregate to County Level

Use spatial joins and aggregation to calculate county-level statistics about vulnerable populations and hospital access.

**Your Task:**
```{r}
#| message: false
#| warning: false
#| results: 'hide'

#county FIPS from tract GEOID
vulnerable_proj <- vulnerable_proj %>%
  mutate(county_fips = substr(GEOID, 1, 5))

#aggregate by county
county_summary <- vulnerable_proj %>%
  st_drop_geometry() %>%
  group_by(county_fips) %>%
  summarise(
    vulnerable_tracts = n(),
    underserved_tracts = sum(underserved),
    pct_underserved = underserved_tracts / vulnerable_tracts,
    avg_distance_miles = mean(nearest_hospital_miles, na.rm = TRUE),
    total_vulnerable_pop = sum(pop_65_plus, na.rm = TRUE),
    .groups = "drop"
  )

#join county names using FIPS column from shapefile
county_summary <- county_summary %>%
  left_join(
    pa_counties %>%
      st_drop_geometry() %>%
      mutate(FIPS_COUNT = sprintf("%05d", as.numeric(FIPS_COUNT))) %>%  # pad to 5 digits
      select(FIPS_COUNT, COUNTY_NAM),
    by = c("county_fips" = "FIPS_COUNT")
  ) %>%
  arrange(desc(pct_underserved))

#top 5 counties by % underserved tracts
head(county_summary, 5)

#counties with most vulnerable population 15+ miles from hospitals
county_population_far <- vulnerable_proj %>%
  filter(underserved) %>%
  st_drop_geometry() %>%
  group_by(county_fips) %>%
  summarise(vulnerable_people_far = sum(pop_65_plus, na.rm = TRUE)) %>%
  left_join(
    pa_counties %>%
      st_drop_geometry() %>%
      mutate(FIPS_COUNT = sprintf("%05d", as.numeric(FIPS_COUNT))) %>%  # pad to 5 digits
      select(FIPS_COUNT, COUNTY_NAM),
    by = c("county_fips" = "FIPS_COUNT")
  ) %>%
  arrange(desc(vulnerable_people_far))

head(county_population_far, 5)

#5 counties highest % underserved vulnerable tracts
head(county_summary, 5) 
#county FIPS 42033, 42053, 42067, 42003, 42007
  #42033 – Clearfield County, PA
  #42053 – Forest County, PA
  #42067 – Juniata County, PA
  #42003 – Allegheny County, PA
  #42007 – Beaver County, PA

# Which counties have the most vulnerable people living far from hospitals?
head(county_population_far, 5)
# county FIPS 42033, 42053, 42067

```

**Required county-level statistics:**
- Number of vulnerable tracts
- Number of underserved tracts  
- Percentage of vulnerable tracts that are underserved
- Average distance to nearest hospital for vulnerable tracts
- Total vulnerable population

**Questions to answer:** 
- Which 5 counties have the highest percentage of underserved vulnerable tracts? 42033/Clearfield County,
42053/Forest County, 
42067/Juniata County, 
42003/Allegheny County, 
42007/Beaver County

- Which counties have the most vulnerable people living far from hospitals? 
42033/Clearfield County (987 people), 
42053/Forest County (802 people), 
42067/Juniata County (492 people)

- Are there any patterns in where underserved counties are located? 
  Aside from Juniata County, 4 of the 5 identified counties are in Western PA. They otherwise have no similar geographic patterns, except that all except for Allegheny County (containing Pittsburgh and its suburbs) may be considered more rural.

---

#### Step 7: Create Summary Table 

Create a professional table showing the top 10 priority counties for healthcare investment.

**Your Task:**
```{r}
#| message: false
#| warning: false
#| results: 'hide'

# Create and format priority counties table
# Map county FIPS codes to actual county names
priority_counties <- county_population_far %>%
  filter(county_fips %in% c("42033", "42053", "42067")) %>%  # only your priority counties
  mutate(
    COUNTY_NAME = case_when(
      county_fips == "42033" ~ "Clearfield County",
      county_fips == "42053" ~ "Forest County",
      county_fips == "42067" ~ "Juniata County"
    )
  ) %>%
  left_join(
    county_summary %>%
      select(county_fips, pct_underserved) %>%
      mutate(pct_underserved = percent(pct_underserved, accuracy = 1)),
    by = "county_fips"
  ) %>%
  mutate(
    vulnerable_people_far = comma(vulnerable_people_far)  # format numbers
  ) %>%
  select(
    COUNTY_NAME,
    vulnerable_people_far,
    pct_underserved
  ) %>%
  arrange(desc(vulnerable_people_far))  # sort by number of vulnerable people far from hospitals

# Create formatted table with kable
knitr::kable(
  priority_counties,
  caption = "Top Pennsylvania Counties for Healthcare Investment: Vulnerable Populations Far from Hospitals",
  col.names = c("County", "Vulnerable People >15 miles from Hospital", "% of Vulnerable Tracts Underserved"),
  align = c("l", "r", "r")
)
```

**Requirements:**
- Use `knitr::kable()` or similar for formatting
- Include descriptive column names
- Format numbers appropriately (commas for population, percentages, etc.)
- Add an informative caption
- Sort by priority (you decide the metric)

---

## Part 2: Comprehensive Visualization 

Using the skills from Week 3 (Data Visualization), create publication-quality maps and charts.

### Map 1: County-Level Choropleth 

Create a choropleth map showing healthcare access challenges at the county level.

**Your Task:**
```{r}
#| message: false
#| warning: false
#| results: 'hide'

# Create county-level access map

#make counties and hospitals same CRS for plot
county_map <- pa_counties %>%
  st_transform(4326) %>%
  left_join(
    county_summary %>%
      mutate(county_fips3 = substr(county_fips, 3, 5)) %>%  # drop the 42
      select(county_fips3, pct_underserved),
    by = c("FIPS_COUNT" = "county_fips3")
  )

hospitals_map <- hospitals %>%
  st_transform(4326)

county_map <- county_map %>%
  mutate(underserved_category = case_when(
    is.na(pct_underserved) ~ "No data",
    pct_underserved == 0 ~ "0%",
    pct_underserved > 0 ~ ">0%"
  ))

#choropleth with hospital points
ggplot() +
  geom_sf(data = county_map, aes(fill = underserved_category), color = "white", size = 0.3) +
  geom_sf(data = hospitals_map, shape = 21, color = "black", fill = "violetred4", size = 2, alpha = 0.7) +
  scale_fill_manual(
    values = c("0%" = "lightblue", ">0%" = "lightskyblue4", "No data" = "grey90"),
    name = "% Vulnerable Tracts Underserved"
  ) +
  theme_void() +
  labs(
    title = "Healthcare Access Challenges in Pennsylvania",
    subtitle = "Counties colored by % of vulnerable tracts >15 miles from nearest hospital",
    caption = "Hospital locations shown as pink points. Vulnerable tracts are 25%+ elderly and low-income (<= $45k) populations."
  )



```

**Requirements:**
- Fill counties by percentage of vulnerable tracts that are underserved
- Include hospital locations as points
- Use an appropriate color scheme
- Include clear title, subtitle, and caption
- Use `theme_void()` or similar clean theme
- Add a legend with formatted labels

---

### Map 2: Detailed Vulnerability Map 

Create a map highlighting underserved vulnerable tracts.

**Your Task:**
```{r}
#| message: false
#| warning: false
#| results: 'hide'

# Create detailed tract-level map

#make vulnerable tracts match plotting CRS WGS84
vulnerable_map <- vulnerable_proj %>%
  st_transform(4326) %>%
  mutate(underserved_category = case_when(
    is.na(underserved) ~ "Not vulnerable",
    underserved == TRUE ~ "Underserved",
    TRUE ~ "Vulnerable"
  ))

#tract-level map
ggplot() +
  geom_sf(data = county_map, fill = NA, color = "grey90", size = 0.3) +
  geom_sf(data = filter(vulnerable_map, underserved_category == "Vulnerable"),
          fill = "lightcyan", color = "grey50", size = 0.1) +
  geom_sf(data = filter(vulnerable_map, underserved_category == "Underserved"),
          fill = "cyan4", color = "darkslategrey", size = 0.2, alpha = 0.7) +
  geom_sf(data = hospitals_map, shape = 21, color = "black", fill = "violetred4", size = 2, alpha = 0.7) +
  labs(
    title = "Underserved Vulnerable Tracts in Pennsylvania",
    subtitle = "Dark blue tracts are >15 miles from nearest hospital; hospitals shown as points",
    caption = "Counties shown as grey outlines. Vulnerable tracts are 25%+ elderly and low-income populations."
  ) +
  theme_void() +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12),
    plot.caption = element_text(size = 10)
  )

```

**Requirements:**
- Show underserved vulnerable tracts in a contrasting color
- Include county boundaries for context
- Show hospital locations
- Use appropriate visual hierarchy (what should stand out?)
- Include informative title and subtitle

---

### Chart: Distribution Analysis

Create a visualization showing the distribution of distances to hospitals for vulnerable populations.

**Your Task:**
```{r}
# Create distribution visualization

#| message: false
#| warning: false
#| results: 'hide'

#histogram with 1 mile bins

ggplot(vulnerable_proj, aes(x = nearest_hospital_miles)) +
  geom_histogram(
    binwidth = 1,                 
    fill = "lightblue",
    color = "grey40",
    alpha = 0.8
  ) +
  geom_vline(
    xintercept = 15,              
    linetype = "dashed",
    color = "magenta3",
    linewidth = 1
  ) +
  labs(
    title = "Distribution of Distances to Nearest Hospital",
    subtitle = "Vulnerable Census Tracts in Pennsylvania",
    x = "Distance to Nearest Hospital (miles)",
    y = "Number of Vulnerable Tracts",
    caption = "Pink dashed line indicates 15-mile threshold after which tracts are considered to be underserved"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 12),
    plot.caption = element_text(size = 10)
  )


```

**Requirements:**
- Clear axes labels with units
- Appropriate title
- Professional formatting
- Brief interpretation (1-2 sentences as a caption or in text)

---

## Part 3: Bring Your Own Data Analysis 

Choose your own additional spatial dataset and conduct a supplementary analysis.

### Challenge Options

---

#### Infrastructure & Services

**Option E: Polling Place Accessibility**
- **Data:** Polling Places, SEPTA stops, Census tracts (elderly population, disability rates)
- **Question:** "Are polling places accessible for elderly and disabled voters?"
- **Operations:** Buffer polling places and transit stops, identify vulnerable populations, find areas lacking access
- **Policy relevance:** Voting rights, election infrastructure, ADA compliance

---

### Data Sources

**OpenDataPhilly:** https://opendataphilly.org/datasets/
- Most datasets available as GeoJSON, Shapefile, or CSV with coordinates
- Always check the Metadata for a data dictionary of the fields.

**Additional Sources:**
- **Pennsylvania Open Data:** https://data.pa.gov/
- **Census Bureau (via tidycensus):** Demographics, economic indicators, commute patterns
- **TIGER/Line (via tigris):** Geographic boundaries

---

### Your Analysis

**Your Task:**

1. **Find and load additional data**
   - Document your data source
   - Check and standardize the CRS
   - Provide basic summary statistics

```{r}
#| message: false
#| warning: false
#| results: 'hide'

# Load your additional dataset
#Infrastructure & Services
pollingplaces <- st_read("C:/Users/macke/Downloads/polling_places.geojson")

#philly census tracts
tracts_phl <- tracts(state = "PA", county = "Philadelphia", year = 2022)

#total population
total_pop <- get_acs(
  geography = "tract",
  variables = "B01003_001",  
  state = "PA",
  county = "Philadelphia",
  year = 2022
) %>%
  select(GEOID, total_population = estimate)

#median hh income
hhincome_data <- get_acs(
  geography = "tract",
  variables = "B19013_001",  # Median household income
  state = "PA",
  county = "Philadelphia",
  year = 2022
) %>%
  select(GEOID, median_income = estimate)

#elderly population 65+ total
elderly_vars <- c(
  "B01001_020", "B01001_021", "B01001_022",
  "B01001_023", "B01001_024", "B01001_025",
  "B01001_044", "B01001_045", "B01001_046",
  "B01001_047", "B01001_048", "B01001_049"
)
elderly_data <- get_acs(
  geography = "tract",
  variables = elderly_vars,
  state = "PA",
  county = "Philadelphia",
  year = 2022
)
elderly_total <- elderly_data %>%
  group_by(GEOID) %>%
  summarise(
    elderly_65_plus = sum(estimate, na.rm = TRUE),
    .groups = "drop"
  )


```

**Questions**
- What dataset did you choose and why?
  Polling places
  
- What is the data source and date?
  OpenDataPhilly, 10/28/2025 (last data update)
  
- How many features does it contain?
  67 features
  
- What CRS is it in? Did you need to transform it?
  WGS84, no transformation needed
  
---

2. **Research question**

Do census tracts with high elderly populations and low median incomes have adequate walking-distance access to polling places? 

---

3. **Spatial analysis**

Use at least TWO spatial operations to answer your research question.

**Required operations (choose 2+):**
- Buffers
- Spatial joins
- Spatial filtering with predicates
- Distance calculations
- Intersections or unions
- Point-in-polygon aggregation

**Your Task:**
```{r}
#| message: false
#| warning: false
#| results: 'hide'

# Your spatial analysis

#join total pop, hh income, elderly
tracts_phl <- tracts_phl %>%
  left_join(total_pop, by = "GEOID") %>%
  left_join(hhincome_data, by = "GEOID") %>%
  left_join(elderly_total, by = "GEOID")

#vulnerable flags #same as parts 1 & 2
tracts_phl <- tracts_phl %>%
  mutate(
    pct_elderly = elderly_65_plus / total_population,
    high_elderly = pct_elderly >= 0.25,
    low_income = median_income <= 45000,
    vulnerable = high_elderly & low_income
  )

#CRS
st_crs(tracts_phl)
st_crs(pollingplaces)
pollingplaces <- st_transform(pollingplaces, st_crs(tracts_phl))

#tract centroids
tract_centroids <- st_centroid(tracts_phl)

#dist matrix in meters 
distance_matrix <- st_distance(tract_centroids, pollingplaces)

#min dist by tract
min_distance_meters <- apply(distance_matrix, 1, min)

#meters to miles
tracts_phl$distance_to_polling_miles <- as.numeric(min_distance_meters) / 1609.34

#underserved
tracts_phl <- tracts_phl %>%
  mutate(
    underserved = distance_to_polling_miles > 0.5,
    underserved_vulnerable = vulnerable & underserved
  )

#summary stats/counts
summary_stats <- tracts_phl %>%
  summarise(
    total_tracts = n(),
    vulnerable_tracts = sum(vulnerable, na.rm = TRUE),
    underserved_tracts = sum(underserved, na.rm = TRUE),
    underserved_vulnerable_tracts = sum(underserved_vulnerable, na.rm = TRUE),
    average_distance_miles = mean(distance_to_polling_miles, na.rm = TRUE),
    max_distance_miles = max(distance_to_polling_miles, na.rm = TRUE)
  )
summary_stats
knitr::kable(
  summary_stats,
  caption = "Summary of Polling Place Accessibility and Vulnerability in Philadelphia",
  align = "c",
  digits = 2
)

#vulnerable? underserved? both?
tracts_phl <- tracts_phl %>%
  mutate(
    access_category = case_when(
      vulnerable & underserved ~ "Vulnerable & Underserved",
      vulnerable & !underserved ~ "Vulnerable Only",
      !vulnerable & underserved ~ "Underserved Only",
      TRUE ~ "Neither"
    )
  )

#map
ggplot() +
  geom_sf(
    data = tracts_phl,
    aes(fill = access_category),
    color = "white",
    size = 0.2
  ) +
  geom_sf(
    data = pollingplaces,
    shape = 21,
    fill = "plum",
    color = "white",
    size = 1.25
  ) +
  scale_fill_manual(
    name = "Access Classification",
    values = c(
      "Vulnerable & Underserved" = "darkred",
      "Vulnerable Only" = "cadetblue3",
      "Underserved Only" = "steelblue",
      "Neither" = "grey90"
    )
  ) +
  labs(
    title = "Polling Place Access and Elderly or Low-Income Vulnerability",
    subtitle = "Philadelphia Census Tracts by Demographic and Spatial Criteria",
    caption = "Underserved defined as >0.5 miles from nearest polling place; Vulnerable defined as 25%+ age 65+ & Median Income $45,000 or less"
  ) +
  theme_void() +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12)
  )
```

**Your interpretation:**
From this analysis, I found no census tracts in Philadelphia county that were both vulnerable and underserved. This suggests that polling place distribution may not disproportionately burden tracts that have both higher-than-state-average elderly populations AND low median incomes. That said, areas which are vulnerable OR underserved still deserve extra attention. And, with additional time for research, I would be interested to see whether areas with ONLY high elderly populations OR low median incomes are ever both vulnerable and underserved.


## Finally - Feedback Incorporation!

In this 2nd lab, I took more creative freedom in deleting unnecessary written elements directly from the assignment; in the first lab, I was hesitant to delete potentially necessary context or instructions, but after reading feedback understand that I can adjust these labs for my own portfolio. Additionally, I tried to make more frequent use of more formal table printing; I'm not sure that I had the issue (at least not frequently) of having just head() or glimpse(), but the reminder certainly made me aware to check! Thanks for your feedback--looking forward to continuing to strengthen my skills!

---
