---
title: "Lab 2: Spatial Analysis and Visualization"
subtitle: "Healthcare Access and Equity in Pennsylvania"
author: "Mackenna Amole"
date: today
format: 
  html:
    code-fold: false
    toc: true
    toc-location: left
    theme: cosmo
    embed-resources: true
execute:
  warning: false
  message: false
---

## Assignment Overview

**Learning Objectives:**
- Apply spatial operations to answer policy-relevant research questions
- Integrate census demographic data with spatial analysis
- Create publication-quality visualizations and maps
- Work with spatial data from multiple sources
- Communicate findings effectively for policy audiences

---

## Part 1: Healthcare Access for Vulnerable Populations

### Research Question

**Which Pennsylvania counties have the highest proportion of vulnerable populations (elderly + low-income) living far from hospitals?**

Your analysis should identify counties that should be priorities for healthcare investment and policy intervention.

### Required Analysis Steps

Complete the following analysis, documenting each step with code and brief explanations:

#### Step 1: Data Collection (5 points)

Load the required spatial data:
- Pennsylvania county boundaries
- Pennsylvania hospitals (from lecture data)
- Pennsylvania census tracts

**Your Task:**
```{r}
# Load required packages
library(sf)
library(tidyverse)
library(tigris)
library(tidycensus)
library(scales)
library(patchwork)
library(here)

# Load spatial data
pa_counties <- st_read(here("labs/lab_2/CPLN-5920-SP26-main/lectures/week_04/data/Pennsylvania_County_Boundaries.shp"))
#pa_counties <- st_read(here("CPLN-5920-SP26-main/CPLN-5920-SP26-main/lectures/week_04/data/Pennsylvania_County_Boundaries.shp"))
#districts <- st_read(here("CPLN-5920-SP26-main/CPLN-5920-SP26-main/lectures/week_04/data/districts.geojson"))
districts <- st_read(here("labs/lab_2/CPLN-5920-SP26-main/lectures/week_04/data/districts.geojson"))
#hospitals <- st_read(here("CPLN-5920-SP26-main/CPLN-5920-SP26-main/lectures/week_04/data/hospitals.geojson"))
hospitals <- st_read(here("labs/lab_2/CPLN-5920-SP26-main/lectures/week_04/data/hospitals.geojson"))
census_tracts <- tracts(state = "PA", cb = TRUE)
metro_areas <- core_based_statistical_areas(cb = TRUE)



# Check that all data loaded correctly
glimpse(pa_counties)
glimpse(districts)
glimpse(hospitals)


# Check coordinates
st_crs(pa_counties)
st_crs(districts)
st_crs(hospitals)
st_crs(metro_areas)
st_crs(census_tracts)

```

**Questions to answer:**
- How many hospitals are in your dataset? 223
- How many census tracts? 3445
- What coordinate reference system is each dataset in? WGS84, except for metro_areas and census_tracts which are in NAD83

---

#### Step 2: Get Demographic Data 

Use `tidycensus` to download tract-level demographic data for Pennsylvania.

**Required variables:**
- Total population
- Median household income
- Population 65 years and over (you may need to sum multiple age categories)

**Your Task:**
```{r}
# Get demographic data from ACS

# Get demographic data from ACS (tract-level, all of PA)
demo_vars <- c(
  total_pop = "B01003_001",
  med_income = "B19013_001",
  male_65_66 = "B01001_020",
  male_67_69 = "B01001_021",
  male_70_74 = "B01001_022",
  male_75_79 = "B01001_023",
  male_80_84 = "B01001_024",
  male_85_plus = "B01001_025",
  female_65_66 = "B01001_044",
  female_67_69 = "B01001_045",
  female_70_74 = "B01001_046",
  female_75_79 = "B01001_047",
  female_80_84 = "B01001_048",
  female_85_plus = "B01001_049"
)
tract_demographics <- get_acs(
  geography = "tract",
  state = "PA",
  variables = demo_vars,
  year = 2022,
  survey = "acs5",
  geometry = TRUE,
  output = "wide"
)
tract_demographics <- tract_demographics %>%
  mutate(
    pop_65_plus =
      male_65_66E + male_67_69E + male_70_74E +
      male_75_79E + male_80_84E + male_85_plusE +
      female_65_66E + female_67_69E + female_70_74E +
      female_75_79E + female_80_84E + female_85_plusE
  )
#tract demographics already has geometry

#tracts with missing income
sum(is.na(tract_demographics$med_incomeE))

#median income PA tracts
median(tract_demographics$med_incomeE, na.rm = TRUE)

```

**Questions to answer:**
- What year of ACS data are you using? 2022
- How many tracts have missing income data? 63
- What is the median income across all PA census tracts? $70,188

---

#### Step 3: Define Vulnerable Populations 

Identify census tracts with vulnerable populations based on TWO criteria:
1. Low median household income (choose an appropriate threshold)
2. Significant elderly population (choose an appropriate threshold)

**Your Task:**
```{r}

#identifying vulnerable: $45k b/c less than median but more than poverty level in PA, 25% b/c PA average is about 20% so this represents a higher concentration
tract_demographics <- tract_demographics %>%
  mutate(
    pct_65_plus = pop_65_plus / total_popE,
    vulnerable = if_else(
      med_incomeE < 45000 & pct_65_plus > 0.25,
      TRUE, FALSE
    )
  )


# Filter for vulnerable tracts based on your criteria

vulnerable_tracts <- tract_demographics %>%
  filter(vulnerable == TRUE)


```

**Questions to answer:**
- What income threshold did you choose and why? I chose 45,000 as the median income threshold. The PA tract average was just slightly over 70,000 and the PA poverty level is approx. 20,000 for 2-person household/approx. 25,000 for 3-person household; 45,000 falls in the middle of this range, highlighting people who are low-income and below state median but not necessarily living below the poverty level.
- What elderly population threshold did you choose and why? I chose 25% age 65+. The PA average population over 65 is about 20% (according to a quick google search for the state scale), so I raised this slightly to identify places with even higher than average concentrations.
- How many tracts meet your vulnerability criteria? 78
- What percentage of PA census tracts are considered vulnerable by your definition? approx. 2.26%

---

#### Step 4: Calculate Distance to Hospitals 

For each vulnerable tract, calculate the distance to the nearest hospital.

**Your Task:**
```{r}
# Transform to appropriate projected CRS
vulnerable_proj <- st_transform(vulnerable_proj, 2272)
hospitals_proj <- st_transform(hospitals, 2272)

#tract centroids
tract_centroids <- st_centroid(vulnerable_proj)

#distance matrix to all hospitals
distance_matrix <- st_distance(tract_centroids, hospitals_proj)

#min dist to nearest hospital in ft
nearest_distance_ft <- apply(distance_matrix, 1, min)

#ft to miles
vulnerable_proj$nearest_hospital_miles <- as.numeric(nearest_distance_ft) / 5280

#avg dist
mean_distance <- mean(vulnerable_proj$nearest_hospital_miles, na.rm = TRUE)
mean_distance

#max dist
max_distance <- max(vulnerable_proj$nearest_hospital_miles, na.rm = TRUE)
max_distance

#number of tracts > 15 miles from hospital
num_far <- sum(vulnerable_proj$nearest_hospital_miles > 15)
num_far

```

**Requirements:**
- Use an appropriate projected coordinate system for Pennsylvania
- Calculate distances in miles
- Explain why you chose your projection

**Questions to answer:**
- What is the average distance to the nearest hospital for vulnerable tracts? approx. 3.88 miles
- What is the maximum distance? 18.13 miles
- How many vulnerable tracts are more than 15 miles from the nearest hospital? 3
- Why did you choose this projection? I transformed the data to Pennsylvania State Plane South (EPSG:2272). This a projected coordinate system designed specifically for Pennsylvania (the south, so chose south because Philadelphia and Pittsburgh are both in the southern half).

---

#### Step 5: Identify Underserved Areas 

Define "underserved" as vulnerable tracts that are more than 15 miles from the nearest hospital.

**Your Task:**
```{r}
# Create underserved variable
vulnerable_proj <- vulnerable_proj %>%
  mutate(underserved = nearest_hospital_miles > 15)

#number underserved tracts
num_underserved <- sum(vulnerable_proj$underserved)
num_underserved

#% vulnerable tracts that are underserved
pct_underserved <- (3/78 *100)
pct_underserved
```

**Questions to answer:** 
- How many tracts are underserved? 3
- What percentage of vulnerable tracts are underserved? 3.85%
- Does this surprise you? Why or why not? This number feels low to me, especially with 78 vulnerably identified tracts. It does seem like a good thing to know that even though vulnerable tracts exist across the state, there are not nearly as many which are a far distance from a hospital.

---

#### Step 6: Aggregate to County Level

Use spatial joins and aggregation to calculate county-level statistics about vulnerable populations and hospital access.

**Your Task:**
```{r}
#county FIPS from tract GEOID
vulnerable_proj <- vulnerable_proj %>%
  mutate(county_fips = substr(GEOID, 1, 5))

#aggregate by county
county_summary <- vulnerable_proj %>%
  st_drop_geometry() %>%
  group_by(county_fips) %>%
  summarise(
    vulnerable_tracts = n(),
    underserved_tracts = sum(underserved),
    pct_underserved = underserved_tracts / vulnerable_tracts,
    avg_distance_miles = mean(nearest_hospital_miles, na.rm = TRUE),
    total_vulnerable_pop = sum(pop_65_plus, na.rm = TRUE),
    .groups = "drop"
  )

#join county names using FIPS column from shapefile
county_summary <- county_summary %>%
  left_join(
    pa_counties_proj %>%
      st_drop_geometry() %>%
      mutate(FIPS_COUNT = sprintf("%05d", as.numeric(FIPS_COUNT))) %>%  # pad to 5 digits
      select(FIPS_COUNT, COUNTY_NAM),
    by = c("county_fips" = "FIPS_COUNT")
  ) %>%
  arrange(desc(pct_underserved))

#top 5 counties by % underserved tracts
head(county_summary, 5)

#counties with most vulnerable population 15+ miles from hospitals
county_population_far <- vulnerable_proj %>%
  filter(underserved) %>%
  st_drop_geometry() %>%
  group_by(county_fips) %>%
  summarise(vulnerable_people_far = sum(pop_65_plus, na.rm = TRUE)) %>%
  left_join(
    pa_counties_proj %>%
      st_drop_geometry() %>%
      mutate(FIPS_COUNT = sprintf("%05d", as.numeric(FIPS_COUNT))) %>%  # pad to 5 digits
      select(FIPS_COUNT, COUNTY_NAM),
    by = c("county_fips" = "FIPS_COUNT")
  ) %>%
  arrange(desc(vulnerable_people_far))

head(county_population_far, 5)

#5 counties highest % underserved vulnerable tracts
head(county_summary, 5) 
#county FIPS 42033, 42053, 42067, 42003, 42007
  #42033 – Clearfield County, PA
  #42053 – Forest County, PA
  #42067 – Juniata County, PA
  #42003 – Allegheny County, PA
  #42007 – Beaver County, PA

# Which counties have the most vulnerable people living far from hospitals?
head(county_population_far, 5)
# county FIPS 42033, 42053, 42067

```

**Required county-level statistics:**
- Number of vulnerable tracts
- Number of underserved tracts  
- Percentage of vulnerable tracts that are underserved
- Average distance to nearest hospital for vulnerable tracts
- Total vulnerable population

**Questions to answer:** 
- Which 5 counties have the highest percentage of underserved vulnerable tracts? 42033/Clearfield County, 42053/Forest County, 42067/Juniata County, 42003/Allegheny County, 42007/Beaver County

- Which counties have the most vulnerable people living far from hospitals? 42033/Clearfield County (987 people), 42053/Forest County (802 people), 42067/Juniata County (492 people)

- Are there any patterns in where underserved counties are located? Aside from Juniata County, 4 of the 5 identified counties are in Western PA. They otherwise have no similar geographic patterns, except that all except for Allegheny County (containing Pittsburgh and its suburbs) may be considered more rural.

---

#### Step 7: Create Summary Table 

Create a professional table showing the top 10 priority counties for healthcare investment.

**Your Task:**
```{r}
# Create and format priority counties table
# Map county FIPS codes to actual county names
priority_counties <- county_population_far %>%
  filter(county_fips %in% c("42033", "42053", "42067")) %>%  # only your priority counties
  mutate(
    COUNTY_NAME = case_when(
      county_fips == "42033" ~ "Clearfield County",
      county_fips == "42053" ~ "Forest County",
      county_fips == "42067" ~ "Juniata County"
    )
  ) %>%
  left_join(
    county_summary %>%
      select(county_fips, pct_underserved) %>%
      mutate(pct_underserved = percent(pct_underserved, accuracy = 1)),
    by = "county_fips"
  ) %>%
  mutate(
    vulnerable_people_far = comma(vulnerable_people_far)  # format numbers
  ) %>%
  select(
    COUNTY_NAME,
    vulnerable_people_far,
    pct_underserved
  ) %>%
  arrange(desc(vulnerable_people_far))  # sort by number of vulnerable people far from hospitals

# Create formatted table with kable
knitr::kable(
  priority_counties,
  caption = "Top Pennsylvania Counties for Healthcare Investment: Vulnerable Populations Far from Hospitals",
  col.names = c("County", "Vulnerable People >15 miles from Hospital", "% of Vulnerable Tracts Underserved"),
  align = c("l", "r", "r")
)
```

**Requirements:**
- Use `knitr::kable()` or similar for formatting
- Include descriptive column names
- Format numbers appropriately (commas for population, percentages, etc.)
- Add an informative caption
- Sort by priority (you decide the metric)

---

## Part 2: Comprehensive Visualization 

Using the skills from Week 3 (Data Visualization), create publication-quality maps and charts.

### Map 1: County-Level Choropleth 

Create a choropleth map showing healthcare access challenges at the county level.

**Your Task:**
```{r}
# Create county-level access map


# Transform counties and hospitals to same CRS for plotting
county_map <- pa_counties_proj %>%
  st_transform(4326) %>%
  left_join(
    county_summary %>%
      mutate(county_fips3 = substr(county_fips, 3, 5)) %>%  # drop the 42
      select(county_fips3, pct_underserved),
    by = c("FIPS_COUNT" = "county_fips3")
  )

hospitals_map <- hospitals %>%
  st_transform(4326)

county_map <- county_map %>%
  mutate(underserved_category = case_when(
    is.na(pct_underserved) ~ "No data",
    pct_underserved == 0 ~ "0%",
    pct_underserved > 0 ~ ">0%"
  ))

# Plot choropleth with hospital points
ggplot() +
  geom_sf(data = county_map, aes(fill = underserved_category), color = "white", size = 0.3) +
  geom_sf(data = hospitals_map, shape = 21, color = "black", fill = "violetred4", size = 2, alpha = 0.7) +
  scale_fill_manual(
    values = c("0%" = "lightblue", ">0%" = "lightskyblue4", "No data" = "grey90"),
    name = "% Vulnerable Tracts Underserved"
  ) +
  theme_void() +
  labs(
    title = "Healthcare Access Challenges in Pennsylvania",
    subtitle = "Counties colored by % of vulnerable tracts >15 miles from nearest hospital",
    caption = "Hospital locations shown as pink points. Vulnerable tracts are 25%+ elderly and low-income (<= $45k) populations."
  )



```

**Requirements:**
- Fill counties by percentage of vulnerable tracts that are underserved
- Include hospital locations as points
- Use an appropriate color scheme
- Include clear title, subtitle, and caption
- Use `theme_void()` or similar clean theme
- Add a legend with formatted labels

---

### Map 2: Detailed Vulnerability Map 

Create a map highlighting underserved vulnerable tracts.

**Your Task:**
```{r}
# Create detailed tract-level map

# Transform vulnerable tracts to match plotting CRS (WGS84)
vulnerable_map <- vulnerable_proj %>%
  st_transform(4326) %>%
  mutate(underserved_category = case_when(
    is.na(underserved) ~ "Not vulnerable",
    underserved == TRUE ~ "Underserved",
    TRUE ~ "Vulnerable"
  ))

# Plot tract-level map
ggplot() +
  # Use your existing county_map for context (light outlines)
  geom_sf(data = county_map, fill = NA, color = "grey90", size = 0.3) +
  
  # Vulnerable tracts that are not underserved (subtle)
  geom_sf(data = filter(vulnerable_map, underserved_category == "Vulnerable"),
          fill = "lightcyan", color = "grey50", size = 0.1) +
  
  # Underserved tracts (highlight)
  geom_sf(data = filter(vulnerable_map, underserved_category == "Underserved"),
          fill = "cyan4", color = "darkslategrey", size = 0.2, alpha = 0.7) +
  
  # Use your existing hospitals_map
  geom_sf(data = hospitals_map, shape = 21, color = "black", fill = "violetred4", size = 2, alpha = 0.7) +
  
  # Titles and caption
  labs(
    title = "Underserved Vulnerable Tracts in Pennsylvania",
    subtitle = "Dark blue tracts are >15 miles from nearest hospital; hospitals shown as points",
    caption = "Counties shown as grey outlines. Vulnerable tracts are 25%+ elderly and low-income populations."
  ) +
  
  # Clean theme
  theme_void() +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12),
    plot.caption = element_text(size = 10)
  )

```

**Requirements:**
- Show underserved vulnerable tracts in a contrasting color
- Include county boundaries for context
- Show hospital locations
- Use appropriate visual hierarchy (what should stand out?)
- Include informative title and subtitle

---

### Chart: Distribution Analysis

Create a visualization showing the distribution of distances to hospitals for vulnerable populations.

**Your Task:**
```{r}
# Create distribution visualization

#histogram with 1 mile bins

ggplot(vulnerable_proj, aes(x = nearest_hospital_miles)) +
  geom_histogram(
    binwidth = 1,                 
    fill = "lightblue",
    color = "grey40",
    alpha = 0.8
  ) +
  geom_vline(
    xintercept = 15,              
    linetype = "dashed",
    color = "magenta3",
    linewidth = 1
  ) +
  labs(
    title = "Distribution of Distances to Nearest Hospital",
    subtitle = "Vulnerable Census Tracts in Pennsylvania",
    x = "Distance to Nearest Hospital (miles)",
    y = "Number of Vulnerable Tracts",
    caption = "Pink dashed line indicates 15-mile threshold after which tracts are considered to be underserved"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 12),
    plot.caption = element_text(size = 10)
  )



```

**Suggested chart types:**
- Histogram or density plot of distances
- Box plot comparing distances across regions
- Bar chart of underserved tracts by county
- Scatter plot of distance vs. vulnerable population size

**Requirements:**
- Clear axes labels with units
- Appropriate title
- Professional formatting
- Brief interpretation (1-2 sentences as a caption or in text)

---

## Part 3: Bring Your Own Data Analysis 

Choose your own additional spatial dataset and conduct a supplementary analysis.

### Challenge Options

Choose ONE of the following challenge exercises, or propose your own research question using OpenDataPhilly data (https://opendataphilly.org/datasets/). 

**Note these are just loose suggestions to spark ideas - follow or make your own as the data permits and as your ideas evolve. This analysis should include bringing in your own dataset, ensuring the projection/CRS of your layers align and are appropriate for the analysis (not lat/long or geodetic coordinate systems). The analysis portion should include some combination of spatial and attribute operations to answer a relatively straightforward question**

---

#### Education & Youth Services

**Option A: Educational Desert Analysis**
- **Data:** Schools, Libraries, Recreation Centers, Census tracts (child population)
- **Question:** "Which neighborhoods lack adequate educational infrastructure for children?"
- **Operations:** Buffer schools/libraries (0.5 mile walking distance), identify coverage gaps, overlay with child population density
- **Policy relevance:** School district planning, library placement, after-school program siting

**Option B: School Safety Zones**
- **Data:** Schools, Crime Incidents, Bike Network
- **Question:** "Are school zones safe for walking/biking, or are they crime hotspots?"
- **Operations:** Buffer schools (1000ft safety zone), spatial join with crime incidents, assess bike infrastructure coverage
- **Policy relevance:** Safe Routes to School programs, crossing guard placement

---

#### Environmental Justice

**Option C: Green Space Equity** 
- **Data:** Parks, Street Trees, Census tracts (race/income demographics)
- **Question:** "Do low-income and minority neighborhoods have equitable access to green space?"
- **Operations:** Buffer parks (10-minute walk = 0.5 mile), calculate tree canopy or park acreage per capita, compare by demographics
- **Policy relevance:** Climate resilience, environmental justice, urban forestry investment
---

#### Public Safety & Justice

**Option D: Crime & Community Resources**
- **Data:** Crime Incidents, Recreation Centers, Libraries, Street Lights
- **Question:** "Are high-crime areas underserved by community resources?"
- **Operations:** Aggregate crime counts to census tracts or neighborhoods, count community resources per area, spatial correlation analysis
- **Policy relevance:** Community investment, violence prevention strategies
---

#### Infrastructure & Services

**Option E: Polling Place Accessibility**
- **Data:** Polling Places, SEPTA stops, Census tracts (elderly population, disability rates)
- **Question:** "Are polling places accessible for elderly and disabled voters?"
- **Operations:** Buffer polling places and transit stops, identify vulnerable populations, find areas lacking access
- **Policy relevance:** Voting rights, election infrastructure, ADA compliance

---

#### Health & Wellness

**Option F: Recreation & Population Health**
- **Data:** Recreation Centers, Playgrounds, Parks, Census tracts (demographics)
- **Question:** "Is lack of recreation access associated with vulnerable populations?"
- **Operations:** Calculate recreation facilities per capita by neighborhood, buffer facilities for walking access, overlay with demographic indicators
- **Policy relevance:** Public health investment, recreation programming, obesity prevention

---

#### Emergency Services

**Option G: EMS Response Coverage**
- **Data:** Fire Stations, EMS stations, Population density, High-rise buildings
- **Question:** "Are population-dense areas adequately covered by emergency services?"
- **Operations:** Create service area buffers (5-minute drive = ~2 miles), assess population coverage, identify gaps in high-density areas
- **Policy relevance:** Emergency preparedness, station siting decisions

---

#### Arts & Culture

**Option H: Cultural Asset Distribution**
- **Data:** Public Art, Museums, Historic sites/markers, Neighborhoods
- **Question:** "Do all neighborhoods have equitable access to cultural amenities?"
- **Operations:** Count cultural assets per neighborhood, normalize by population, compare distribution across demographic groups
- **Policy relevance:** Cultural equity, tourism, quality of life, neighborhood identity

---

### Data Sources

**OpenDataPhilly:** https://opendataphilly.org/datasets/
- Most datasets available as GeoJSON, Shapefile, or CSV with coordinates
- Always check the Metadata for a data dictionary of the fields.

**Additional Sources:**
- **Pennsylvania Open Data:** https://data.pa.gov/
- **Census Bureau (via tidycensus):** Demographics, economic indicators, commute patterns
- **TIGER/Line (via tigris):** Geographic boundaries

### Recommended Starting Points

**If you're feeling confident:** Choose an advanced challenge with multiple data layers. 
**If you are a beginner, choose something more manageable that helps you understand the basics**


**If you have a different idea:** Propose your own question! Just make sure:
- You can access the spatial data
- You can perform at least 2 spatial operations

### Your Analysis

**Your Task:**

1. **Find and load additional data**
   - Document your data source
   - Check and standardize the CRS
   - Provide basic summary statistics

```{r}
# Load your additional dataset




```

**Questions to answer:**
- What dataset did you choose and why?
- What is the data source and date?
- How many features does it contain?
- What CRS is it in? Did you need to transform it?

---

2. **Pose a research question**

Write a clear research statement that your analysis will answer.

**Examples:**
- "Do vulnerable tracts have adequate public transit access to hospitals?"
- "Are EMS stations appropriately located near vulnerable populations?"
- "Do areas with low vehicle access have worse hospital access?"

---

3. **Conduct spatial analysis**

Use at least TWO spatial operations to answer your research question.

**Required operations (choose 2+):**
- Buffers
- Spatial joins
- Spatial filtering with predicates
- Distance calculations
- Intersections or unions
- Point-in-polygon aggregation

**Your Task:**
```{r}
# Your spatial analysis










```

**Analysis requirements:**
- Clear code comments explaining each step
- Appropriate CRS transformations
- Summary statistics or counts
- At least one map showing your findings
- Brief interpretation of results (3-5 sentences)

**Your interpretation:**

[Write your findings here]


## Finally - A few comments about your incorporation of feedback!
Take a few moments to clean up your markdown document and then write a line or two or three about how you may have incorporated feedback that you recieved after your first assignment. 

---

## Submission Requirements

**What to submit:**

1. **Rendered HTML document posted to your course portfolio** with all code, outputs, maps, and text
   - Use `embed-resources: true` in YAML so it's a single file
   - All code should run without errors
   - All maps and charts should display correctly
2. Submit the correct and working links of your assignment on Canvas

---

