---
title: "Lab 4: Spatial Predictive Analysis"
subtitle: "Analyzing 311 Violations"
author: "Mackenna Amole"
date: 9 April 2026
format: 
  html:
    code-fold: show
    toc: true
    toc-location: left
    theme: cosmo
    embed-resources: true
execute:
  warning: false
  message: false
---

## Assignment Overview

In this lab, I applied spatial predictive modeling techniques demonstrated in an in-class exercise using 311 Sanitation Violations as the predictor variable. I built a complete spatial predictive model, documented my process, and interpreted my results.

### Learning Objectives

-   Adapt example code to analyze a new dataset
-   Build spatial features for predictive modeling
-   Apply count regression techniques to spatial data
-   Implement spatial cross-validation
-   Interpret and communicate model results
-   Critically evaluate model performance

------------------------------------------------------------------------

## Choosing My 311 Violation Type

### Getting the Data

Visit the Chicago 311 Service Requests dataset:

<https://data.cityofchicago.org/stories/s/311-Dataset-Changes-12-11-2018/d7nq-5g7t>

Selection: **SANITATION VIOLAITONS**

------------------------------------------------------------------------

## Completing the Analysis

```{r}
#| message: false
#| warning: false
#| results: 'hide'

#install - comment out later
install.packages("spatstat.explore")
install.packages("spatstat.geom")
install.packages("viridis")

# Load required packages
library(tidyverse)
library(sf)
library(here)
library(viridis)
library(terra)
library(spdep)
library(FNN)
library(MASS)
library(patchwork)
library(knitr)
library(kableExtra)
library(classInt)
# Spatstat for analyzing point patterns - split into sub-packages
library(spatstat.geom)
library(spatstat.explore)


# Set options
options(scipen = 999)  # No scientific notation
set.seed(5080)         # Reproducibility, could be any number

# Create consistent theme for visualizations
theme_crime <- function(base_size = 11) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.title = element_text(face = "bold", size = base_size + 1),
      plot.subtitle = element_text(color = "gray30", size = base_size - 1),
      legend.position = "right",
      panel.grid.minor = element_blank(),
      axis.text = element_blank(),
      axis.title = element_blank()
    )
}

# Set as default
theme_set(theme_crime())

cat("✓ All packages loaded successfully!\n")
cat("✓ Working directory:", getwd(), "\n")

```

#### 1: Data Loading & Exploration

**Analysis** - What you are doing in that step / **Why** this step is important for the analysis / **What** you found or learned from the results

-   1.1: Load 311 data and Chicago spatial boundaries

```{r load-boundaries}
#| message: false

# Load police districts (used for spatial cross-validation)
policeDistricts <- 
  st_read("https://data.cityofchicago.org/api/geospatial/24zt-jpfn?method=export&format=GeoJSON") %>%
  st_transform('ESRI:102271') %>%
  dplyr::select(District = dist_num)

# Load police beats (smaller administrative units)
policeBeats <- 
  st_read("https://data.cityofchicago.org/api/geospatial/n9it-hstw?method=export&format=GeoJSON") %>%
  st_transform('ESRI:102271') %>%
  dplyr::select(Beat = beat_num)

# Load Chicago boundary
chicagoBoundary <- 
  st_read("https://raw.githubusercontent.com/urbanSpatial/Public-Policy-Analytics-Landing/master/DATA/Chapter5/chicagoBoundary.geojson") %>%
  st_transform('ESRI:102271')

cat("✓ Loaded spatial boundaries\n")
cat("  - Police districts:", nrow(policeDistricts), "\n")
cat("  - Police beats:", nrow(policeBeats), "\n")
```

```{r load-sanitation}
sanitation <- read_csv(here("labs/lab_4//311_Sanitation.csv")) %>%
  filter(!is.na(Longitude), !is.na(Latitude)) %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326) %>%
  st_transform('ESRI:102271')

glimpse(sanitation)
cat("Number of sanitation requests:", nrow(sanitation), "\n")
```

-   1.2: Create visualizations showing the spatial distribution of violation type

```{r}
#| message: false
#| warning: false
#| results: 'hide'

# Point map
p1 <- ggplot() +
  geom_sf(data = chicagoBoundary, fill = "gray95", color = "gray60") +
  geom_sf(data = sanitation, color = "#1b9e77", size = 0.1, alpha = 0.4) +
  labs(
    title = "Sanitation 311 Requests",
    subtitle = paste0("Chicago, n = ", nrow(sanitation))
  )

# KDE surface
sanitation_ppp <- as.ppp(
  st_coordinates(sanitation),
  W = as.owin(st_bbox(chicagoBoundary))
)

kde_sanitation <- density.ppp(sanitation_ppp, sigma = 1000, edge = TRUE)
kde_raster <- rast(kde_sanitation)

# Convert to dataframe for plotting
kde_df <- as.data.frame(kde_raster, xy = TRUE)

p2 <- ggplot(kde_df) +
  geom_raster(aes(x = x, y = y, fill = layer)) +
  scale_fill_viridis_c() +
  labs(title = "KDE Surface of Sanitation Requests")

p1 + p2

```

-   1.3: Describe patterns observed

Then write something like:

Sanitation requests are clustered, not evenly distributed
Higher concentrations appear in denser urban areas
KDE shows smooth hot spots that suggest spatial clustering


#### 2: Fishnet Grid Creation

**Analysis** - What you are doing in that step / **Why** this step is important for the analysis / **What** you found or learned from the results

-   2.1: Create a 500m x 500m fishnet grid
-   2.2: Aggregate your violations to grid cells
-   2.3: Visualize the count distribution

```{r}
#| message: false
#| warning: false
#| results: 'hide'

# 2.1 Create fishnet
fishnet <- st_make_grid(chicagoBoundary, cellsize = 500, square = TRUE) %>%
  st_sf() %>%
  mutate(uniqueID = row_number()) %>%
  st_transform('ESRI:102271')

fishnet <- fishnet[chicagoBoundary, ]

# 2.2 Aggregate sanitation
sanitation_fishnet <- st_join(sanitation, fishnet, join = st_within) %>%
  st_drop_geometry() %>%
  group_by(uniqueID) %>%
  summarize(countSanitation = n())

# Join back
fishnet <- fishnet %>%
  left_join(sanitation_fishnet, by = "uniqueID") %>%
  mutate(countSanitation = replace_na(countSanitation, 0))

# 2.3 Map
ggplot() +
  geom_sf(data = fishnet, aes(fill = countSanitation), color = NA) +
  geom_sf(data = chicagoBoundary, fill = NA, color = "white") +
  scale_fill_viridis_c(trans = "sqrt") +
  labs(title = "Sanitation Requests by Grid Cell")

```

#### 3: Spatial Features

**Analysis** - What you are doing in that step / **Why** this step is important for the analysis / **What** you found or learned from the results

-   3.1: Calculate k-nearest neighbor features
-   3.2: Perform Local Moran's I analysis
-   3.3: Identify hot spots and cold spots
-   3.4: Create distance-to-hotspot measures
-   3.5: Join any additional contextual data if you are looking for more to do and really get into this (e.g., demographics, land use)

```{r}
#| message: false
#| warning: false
#| results: 'hide'

# 3.1 KNN
fishnet_coords <- st_coordinates(st_centroid(fishnet))
sanitation_coords <- st_coordinates(sanitation)

nn <- get.knnx(sanitation_coords, fishnet_coords, k = 3)

fishnet <- fishnet %>%
  mutate(sanitation.nn = rowMeans(nn$nn.dist))

# 3.2 Moran's I
coords <- st_coordinates(st_centroid(fishnet))
neighbors <- spdep::knn2nb(spdep::knearneigh(coords, k = 5))
weights <- spdep::nb2listw(neighbors, style = "W")

local_moran <- spdep::localmoran(fishnet$countSanitation, weights)

fishnet <- fishnet %>%
  mutate(
    local_i = local_moran[,1],
    p_value = local_moran[,5],
    moran_class = case_when(
      p_value > 0.05 ~ "Not Significant",
      local_i > 0 & countSanitation > mean(countSanitation) ~ "High-High",
      local_i > 0 ~ "Low-Low",
      local_i < 0 & countSanitation > mean(countSanitation) ~ "High-Low",
      TRUE ~ "Low-High"
    )
  )

# 3.3 Map clusters
ggplot() +
  geom_sf(aes(fill = moran_class), data = fishnet, color = NA)

# 3.4 Distance to hotspots
hotspots <- fishnet %>% filter(moran_class == "High-High")

fishnet <- fishnet %>%
  mutate(
    dist_to_hotspot = as.numeric(
      st_distance(st_centroid(fishnet), st_union(st_centroid(hotspots)))
    )
  )

# 3.5 Join districts
fishnet <- st_join(fishnet, policeDistricts) %>%
  filter(!is.na(District))

```

#### 4: Count Regression Models

**Analysis** - What you are doing in that step / **Why** this step is important for the analysis / **What** you found or learned from the results

-   4.1: Fit Poisson regression
-   4.2: Fit Negative Binomial regression
-   4.3: Compare model fit (AIC)

```{r}
#| message: false
#| warning: false
#| results: 'hide'

fishnet_model <- fishnet %>%
  st_drop_geometry() %>%
  dplyr::select(
    uniqueID,
    District,
    countSanitation,
    sanitation.nn,
    dist_to_hotspot
  ) %>%  
  na.omit()

# 4.1  Poisson
model_poisson <- glm(
  countSanitation ~ sanitation.nn + dist_to_hotspot,
  data = fishnet_model,
  family = "poisson"
)

# Overdispersion
dispersion <- sum(residuals(model_poisson, type = "pearson")^2) / model_poisson$df.residual

# Negative Binomial
model_nb <- glm.nb(
  countSanitation ~ sanitation.nn + dist_to_hotspot,
  data = fishnet_model
)

# Compare
AIC(model_poisson, model_nb)

```

#### 5: Spatial Cross-Validation (2017)

**Analysis** - What you are doing in that step / **Why** this step is important for the analysis / **What** you found or learned from the results

-   5.1: Implement Leave-One-Group-Out cross-validation on 2017 data
-   5.2: Calculate and report error metrics (MAE, RMSE)

```{r}
#| message: false
#| warning: false
#| results: 'hide'

districts <- unique(fishnet_model$District)
cv_results <- data.frame()

for (d in districts) {

  train <- fishnet_model %>% filter(District != d)
  test <- fishnet_model %>% filter(District == d)

  model <- glm.nb(
    countSanitation ~ sanitation.nn + dist_to_hotspot,
    data = train
  )

  preds <- predict(model, test, type = "response")

  mae <- mean(abs(test$countSanitation - preds))
  rmse <- sqrt(mean((test$countSanitation - preds)^2))

  cv_results <- rbind(cv_results,
    data.frame(District = d, MAE = mae, RMSE = rmse))
}

mean(cv_results$MAE)
mean(cv_results$RMSE)

```

#### 6: Model Evaluation

**Analysis** - What you are doing in that step / **Why** this step is important for the analysis / **What** you found or learned from the results

-   6.1: Compare to KDE baseline

```{r}
#| message: false
#| warning: false
#| results: 'hide'

# Add model predictions
fishnet <- fishnet %>%
  mutate(pred_nb = predict(model_nb, fishnet_model, type = "response"))

# Normalize KDE
kde_vals <- terra::extract(kde_raster, vect(fishnet), fun = mean)[,2]

fishnet$kde <- kde_vals

fishnet <- fishnet %>%
  mutate(pred_kde = kde / sum(kde) * sum(countSanitation))

# Compare maps
p1 <- ggplot() + geom_sf(aes(fill = countSanitation), data = fishnet)
p2 <- ggplot() + geom_sf(aes(fill = pred_nb), data = fishnet)
p3 <- ggplot() + geom_sf(aes(fill = pred_kde), data = fishnet)

p1 + p2 + p3

# Metrics
mean(abs(fishnet$countSanitation - fishnet$pred_nb))
mean(abs(fishnet$countSanitation - fishnet$pred_kde))

```

------------------------------------------------------------------------

## Step 3: Write Your Analysis

### Critical Requirement: Explain Each Step

**For each major section, you must explain in your own words:**

-   **What** you are doing in that step
-   **Why** this step is important for the analysis
-   **What** you found or learned from the results

Do not simply copy text from the example or from your AI friend. Think about the purpose of each technique and articulate it in your own words.

------------------------------------------------------------------------

## Step 4: Format Your Document

### Formatting Requirements

Your rendered HTML document should be **professional and easy to read**:

#### ✓ Clean Code

-   Remove unnecessary code, comments, or debugging lines
-   Keep only essential code chunks
-   Use `code-fold: show` in your YAML header

#### ✓ Clear Structure

-   Use headers to organize sections
-   Include a table of contents
-   Add your name and date

#### ✓ Readable Text

-   Ensure all text renders properly (no broken markdown)
-   Check that headers appear as headers (not plain text)
-   Use proper markdown formatting

#### ✓ Quality Visualizations

-   All plots should have titles and labels
-   Maps should be readable
-   Use consistent color schemes

#### ✓ Professional Presentation

-   Proofread for typos
-   Remove any "Your answer here" placeholders
-   Make sure all code runs without errors
-   Make IT NEAT. WE GOTTA LOOK GOOD HERE.

------------------------------------------------------------------------

## Submission Checklist

Before you submit, verify that your document includes:

### Required Components

-   [ ] **Introduction**: Brief description of your chosen 311 violation type and why you selected it
-   [ ] **Data Exploration**: Maps and summary statistics of your violation data
-   [ ] **Fishnet Grid**: Successfully created and visualized grid with aggregated counts
-   [ ] **Spatial Features**: At least 2-3 spatial features calculated (k-NN, distance measures, etc.)
-   [ ] **Local Moran's I**: Spatial autocorrelation analysis with interpretation
-   [ ] **Count Regression**: Both Poisson and Negative Binomial models with comparison
-   [ ] **Spatial Cross-Validation**: LOGO CV implemented on 2017 data with reported metrics
-   [ ] **Temporal Validation**: 2018 data aggregated and tested with 2017 model(OPTIONAL.)
-   [ ] **Validation Comparison**: Comparison of spatial vs. temporal validation performance
-   [ ] **Model Comparison**: Your model vs. KDE baseline for 2018 predictions
-   [ ] **Error Analysis**: Maps and discussion of where model performs well/poorly
-   [ ] **Written Explanations**: Your own words explaining each step

### Technical Requirements

-   [ ] Document renders to HTML without errors
-   [ ] Code is clean and well-organized
-   [ ] All visualizations display properly
-   [ ] Headers and formatting render correctly
-   [ ] File is named successfully posted to your website

------------------------------------------------------------------------
