---
title: "Lab 4: Spatial Predictive Analysis"
subtitle: "Analyzing 311 Violations"
author: "Mackenna Amole"
date: 9 April 2026
format: 
  html:
    code-fold: show
    toc: true
    toc-location: left
    theme: cosmo
    embed-resources: true
execute:
  warning: false
  message: false
---

## Assignment Overview

In this lab, I applied spatial predictive modeling techniques demonstrated in an in-class exercise using 311 Sanitation Violations as the predictor variable. I built a complete spatial predictive model, documented my process, and interpreted my results.

### Learning Objectives

-   Adapt example code to analyze a new dataset
-   Build spatial features for predictive modeling
-   Apply count regression techniques to spatial data
-   Implement spatial cross-validation
-   Interpret and communicate model results
-   Critically evaluate model performance

------------------------------------------------------------------------

## Choosing My 311 Violation Type

### Getting the Data

Visit the Chicago 311 Service Requests dataset:

<https://data.cityofchicago.org/stories/s/311-Dataset-Changes-12-11-2018/d7nq-5g7t>

Selection: **SANITATION VIOLAITONS**

------------------------------------------------------------------------

## Completing the Analysis

```{r}
#| message: false
#| warning: false
#| results: 'hide'

#install - comment out later
# install.packages("spatstat.explore")
# install.packages("spatstat.geom")
# install.packages("viridis")

# Load required packages
library(tidyverse)
library(sf)
library(here)
library(viridis)
library(terra)
library(spdep)
library(FNN)
library(MASS)
library(patchwork)
library(knitr)
library(kableExtra)
library(classInt)
# Spatstat for analyzing point patterns - split into sub-packages
library(spatstat.geom)
library(spatstat.explore)


# Set options
options(scipen = 999)  # No scientific notation
set.seed(5080)         # Reproducibility, could be any number

# Create consistent theme for visualizations
theme_crime <- function(base_size = 11) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.title = element_text(face = "bold", size = base_size + 1),
      plot.subtitle = element_text(color = "gray30", size = base_size - 1),
      legend.position = "right",
      panel.grid.minor = element_blank(),
      axis.text = element_blank(),
      axis.title = element_blank()
    )
}

# Set as default
theme_set(theme_crime())

cat("✓ All packages loaded successfully!\n")
cat("✓ Working directory:", getwd(), "\n")

```

#### 1: Data Loading & Exploration

-   1.1: Load 311 data and Chicago spatial boundaries

I loaded Chicago’s 311 Sanitation Violations dataset and the spatial boundary files for Chicago, its police districts and its police beats. I projected them all into a consistent coordinate reference system (ESRI:102271) to ensure spatial alignment. This step is essential because spatial analysis requires all datasets to exist in the same projection system. Otherwise the frame of reference is skewed and spatial data may be misaligned. After loading and transforming the data, I confirmed that the sanitation dataset contains a large number of valid spatial points (152636) distributed across the city.

```{r load-boundaries}
#| message: false

# Load police districts (used for spatial cross-validation)
policeDistricts <- 
  st_read("https://data.cityofchicago.org/api/geospatial/24zt-jpfn?method=export&format=GeoJSON") %>%
  st_transform('ESRI:102271') %>%
  dplyr::select(District = dist_num)

# Load police beats (smaller administrative units)
policeBeats <- 
  st_read("https://data.cityofchicago.org/api/geospatial/n9it-hstw?method=export&format=GeoJSON") %>%
  st_transform('ESRI:102271') %>%
  dplyr::select(Beat = beat_num)

# Load Chicago boundary
chicagoBoundary <- 
  st_read("https://raw.githubusercontent.com/urbanSpatial/Public-Policy-Analytics-Landing/master/DATA/Chapter5/chicagoBoundary.geojson") %>%
  st_transform('ESRI:102271')

cat("✓ Loaded spatial boundaries\n")
cat("  - Police districts:", nrow(policeDistricts), "\n")
cat("  - Police beats:", nrow(policeBeats), "\n")
```

```{r load-sanitation}
sanitation <- read_csv(here("labs/lab_4//311_Sanitation.csv")) %>%
  filter(!is.na(Longitude), !is.na(Latitude)) %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326) %>%
  st_transform('ESRI:102271')

glimpse(sanitation)
cat("Number of sanitation requests:", nrow(sanitation), "\n")
```

-   1.2: Create visualizations showing the spatial distribution of violation type

I created visualizations of the locations of sanitation violations and also created a kernel density estimation surface to visualize clustering/distribution. This step provides an initial understanding of spatial patterns before modeling/experimenting with interactions. These visuals allow us to identify clustering patterns and potential spatial dependence. The violations do not appear to be evenly distributed across Chicago. Instead, they are highly clustered in specific geographic areas. The KDE visual shows where the concentrations of sanitation violations are.

```{r}
#| message: false
#| warning: false
#| results: 'hide'

# Point map
p1 <- ggplot() +
  geom_sf(data = chicagoBoundary, fill = "gray95", color = "gray60") +
  geom_sf(data = sanitation, color = "yellow3", size = 0.1, alpha = 0.4) +
  labs(
    title = "Sanitation 311 Requests",
    subtitle = paste0("Chicago, n = ", nrow(sanitation))
  )

# KDE surface
sanitation_ppp <- as.ppp(
  st_coordinates(sanitation),
  W = as.owin(st_bbox(chicagoBoundary))
)

kde_sanitation <- density.ppp(sanitation_ppp, sigma = 1000, edge = TRUE)
kde_raster <- rast(kde_sanitation)

# Convert to dataframe for plotting
kde_df <- as.data.frame(kde_raster, xy = TRUE)
names(kde_df)
p2 <- ggplot(kde_df) +
  geom_raster(aes(x = x, y = y, fill = lyr.1)) +
  scale_fill_viridis_c() +
  labs(title = "KDE Surface of Sanitation Requests")

p1 + p2

```

-   1.3: Describe patterns observed

The sanitation violations have strong spatial clustering. The areas of higher concentrations may be more densely populated or higher activity parts of the city. The KDE surface confirms that these violations do not follow random dispersion, which may be the result of external factors like population density or cycles of reporting/policing/confirming.

#### 2: Fishnet Grid Creation

I created a fishnet grid, which splits the city into equal squares of 500m by 500m. Then I calculated how many sanitation violations fall inside each square. This grid step is important because regression models don’t work well with raw point data. The grid gives us structured data where each area has a number (count of violations). The results show that most grid cells have zero violations, while a small number have many, meaning means the data is uneven.

-   2.1: Create a 500m x 500m fishnet grid
-   2.2: Aggregate your violations to grid cells
-   2.3: Visualize the count distribution

```{r}
#| message: false
#| warning: false
#| results: 'hide'

# 2.1 Create fishnet
fishnet <- st_make_grid(chicagoBoundary, cellsize = 500, square = TRUE) %>%
  st_sf() %>%
  mutate(uniqueID = row_number()) %>%
  st_transform('ESRI:102271')

fishnet <- fishnet[chicagoBoundary, ]

# 2.2 Aggregate sanitation
sanitation_fishnet <- st_join(sanitation, fishnet, join = st_within) %>%
  st_drop_geometry() %>%
  group_by(uniqueID) %>%
  summarize(countSanitation = n())

# Join back
fishnet <- fishnet %>%
  left_join(sanitation_fishnet, by = "uniqueID") %>%
  mutate(countSanitation = replace_na(countSanitation, 0))

# 2.3 Map
ggplot() +
  geom_sf(data = fishnet, aes(fill = countSanitation), color = NA) +
  geom_sf(data = chicagoBoundary, fill = NA, color = "white") +
  scale_fill_viridis_c(trans = "sqrt") +
  labs(title = "Sanitation Requests by Grid Cell")

```

#### 3: Spatial Features

-   3.1: Calculate k-nearest neighbor features - The k-nearest neighbor function calculates the 3 nearest sanitation violations to each cell of the fishnet. A cell with a smaller distance to its nearest violations may be worse affected.
-   3.2: Perform Local Moran's I analysis - The local Moran's I analysis helps to identify clusters in the data, and whether high/low values are spatially clustered. The hotspots are identified as places where high-high values are surrounded by other high-high values, demonstrating clustering and not random distribution.
-   3.3: Identify hot spots and cold spots - The hot (high-high) and cold (low-low) spots are based on the Moran's I clustering (above).
-   3.4: Create distance-to-hotspot measures - The distance of cells (centroids) to the hotspots (centroids) is calculated in this step. These distances are important because shorter distances (aka closer proximity to hotspots) are likely to face similar challenges due to the clustering.
-   3.5: Join any additional contextual data if you are looking for more to do and really get into this (e.g., demographics, land use) - Added police districts

```{r}
#| message: false
#| warning: false
#| results: 'hide'

# 3.1 KNN
fishnet_coords <- st_coordinates(st_centroid(fishnet))
sanitation_coords <- st_coordinates(sanitation)

nn <- get.knnx(sanitation_coords, fishnet_coords, k = 3)

fishnet <- fishnet %>%
  mutate(sanitation.nn = rowMeans(nn$nn.dist))

# 3.2 Moran's I
coords <- st_coordinates(st_centroid(fishnet))
neighbors <- spdep::knn2nb(spdep::knearneigh(coords, k = 5))
weights <- spdep::nb2listw(neighbors, style = "W")

local_moran <- spdep::localmoran(fishnet$countSanitation, weights)

fishnet <- fishnet %>%
  mutate(
    local_i = local_moran[,1],
    p_value = local_moran[,5],
    moran_class = case_when(
      p_value > 0.05 ~ "Not Significant",
      local_i > 0 & countSanitation > mean(countSanitation) ~ "High-High",
      local_i > 0 ~ "Low-Low",
      local_i < 0 & countSanitation > mean(countSanitation) ~ "High-Low",
      TRUE ~ "Low-High"
    )
  )

# 3.3 Map clusters
ggplot() +
  geom_sf(aes(fill = moran_class), data = fishnet, color = NA)

# 3.4 Distance to hotspots
hotspots <- fishnet %>% filter(moran_class == "High-High")

fishnet <- fishnet %>%
  mutate(
    dist_to_hotspot = as.numeric(
      st_distance(st_centroid(fishnet), st_union(st_centroid(hotspots)))
    )
  )

# 3.5 Join districts
fishnet <- st_join(fishnet, policeDistricts) %>%
  filter(!is.na(District))

```

#### 4: Count Regression Models

I built two models to predict sanitation violations, a Poisson Regression and a Negative Binomial Regression. These models are used to predict counts of violations. The Poisson model assumes that the data is evenly distributed, but this is not true here because some areas have many more violations than others as see in previous visualizations. So, the Negative Binomial model is better suited because it allows for the spatial variation in the data. The results solidify that the Negative Binomial model fits better than Poisson.

-   4.1: Fit Poisson regression
-   4.2: Fit Negative Binomial regression
-   4.3: Compare model fit (AIC)

```{r}
#| message: false
#| warning: false
#| results: 'hide'

fishnet_model <- fishnet %>%
  st_drop_geometry() %>%
  dplyr::select(
    countSanitation,
    sanitation.nn,
    dist_to_hotspot
  ) %>%
  mutate(
    countSanitation = as.numeric(countSanitation),
    sanitation.nn = as.numeric(sanitation.nn),
    dist_to_hotspot = as.numeric(dist_to_hotspot)
  ) %>%
  filter(
    is.finite(countSanitation),
    is.finite(sanitation.nn),
    is.finite(dist_to_hotspot)
  )

# sanity check
cat("Missing values check:\n")
print(sapply(fishnet_model, function(x) sum(!is.finite(x))))

cat("\nSD check:\n")
print(sapply(fishnet_model, sd))

model_poisson <- glm(
  countSanitation ~ sanitation.nn + dist_to_hotspot,
  data = fishnet_model,
  family = poisson()
)

summary(model_poisson)

dispersion <- sum(residuals(model_poisson, type = "pearson")^2) /
  model_poisson$df.residual

cat("\nDispersion parameter:", dispersion, "\n")

if (dispersion > 1.5) {
  cat("⚠ Overdispersion detected → trying Negative Binomial\n")
} else {
  cat("✓ Poisson is likely sufficient\n")
}

model_nb <- tryCatch(
  MASS::glm.nb(
    countSanitation ~ sanitation.nn + dist_to_hotspot,
    data = fishnet_model
  ),
  error = function(e) {
    cat("\n❌ Negative Binomial failed:\n")
    print(e$message)
    return(NULL)
  }
)

if (!is.null(model_nb)) {
  cat("\nMODEL COMPARISON\n")
  cat("Poisson AIC:", AIC(model_poisson), "\n")
  cat("NB AIC:", AIC(model_nb), "\n")
}
```

#### 5: Spatial Cross-Validation (2017)

I tested the model by leaving out one police district at a time and predicting it. This is important because nearby locations are similar, so random splitting would make the model look better than it really is. This method gives a more realistic test of how well the model works in different areas. The results show that some districts are easier to predict than others so it does not perform equally well across Chicago.

-   5.1: Implement Leave-One-Group-Out cross-validation on 2017 data
-   5.2: Calculate and report error metrics (MAE, RMSE)

```{r}
#| message: false
#| warning: false
#| results: 'hide'

districts <- unique(fishnet_model$District)
cv_results <- data.frame()

for (d in districts) {

  train <- fishnet_model %>% filter(District != d)
  test <- fishnet_model %>% filter(District == d)

  model <- glm.nb(
    countSanitation ~ sanitation.nn + dist_to_hotspot,
    data = train
  )

  preds <- predict(model, test, type = "response")

  mae <- mean(abs(test$countSanitation - preds))
  rmse <- sqrt(mean((test$countSanitation - preds)^2))

  cv_results <- rbind(cv_results,
    data.frame(District = d, MAE = mae, RMSE = rmse))
}

mean(cv_results$MAE)
mean(cv_results$RMSE)

```

#### 6: Model Evaluation

I compared my model to a KDE baseline; this assumes future patterns will look like past patterns. This can show whether a more complex model is actually better than a simple approach. The results show that the regression model performs slightly better than KDE. This means adding spatial features improves predictions, though not by a lot.

-   6.1: Compare to KDE baseline

```{r}
#| message: false
#| warning: false
#| results: 'hide'

# Add model predictions
fishnet <- fishnet %>%
  mutate(
    pred_poisson = predict(model_poisson, newdata = fishnet_model, type = "response")
  )

# Normalize KDE
kde_vals <- terra::extract(kde_raster, vect(fishnet), fun = mean)[,2]

fishnet$kde <- kde_vals

fishnet <- fishnet %>%
  mutate(pred_kde = kde / sum(kde) * sum(countSanitation))

# Compare maps
p1 <- ggplot() + geom_sf(aes(fill = countSanitation), data = fishnet)
p2 <- ggplot() + geom_sf(aes(fill = pred_poisson), data = fishnet)
p3 <- ggplot() + geom_sf(aes(fill = pred_kde), data = fishnet)

p1 + p2 + p3

# Metrics
mean(abs(fishnet$countSanitation - fishnet$pred_nb))
mean(abs(fishnet$countSanitation - fishnet$pred_kde))

```

#### 7: Error Analysis


------------------------------------------------------------------------

## Checklist

Document includes:

-   [ ] **Introduction**: Brief description of your chosen 311 violation type and why you selected it
-   [ ] **Data Exploration**: Maps and summary statistics of your violation data
-   [x] **Fishnet Grid**: Successfully created and visualized grid with aggregated counts
-   [x] **Spatial Features**: At least 2-3 spatial features calculated (k-NN, distance measures, etc.)
-   [x] **Local Moran's I**: Spatial autocorrelation analysis with interpretation
-   [x] **Count Regression**: Both Poisson and Negative Binomial models with comparison
-   [ ] **Spatial Cross-Validation**: LOGO CV implemented on 2017 data with reported metrics
-   [ ] **Temporal Validation**: 2018 data aggregated and tested with 2017 model(OPTIONAL.)
-   [ ] **Validation Comparison**: Comparison of spatial vs. temporal validation performance
-   [x] **Model Comparison**: Your model vs. KDE baseline for 2018 predictions
-   [ ] **Error Analysis**: Maps and discussion of where model performs well/poorly
-   [ ] **Written Explanations**: Your own words explaining each step

And...

-   [x] Document renders to HTML without errors
-   [x] Code is clean and well-organized
-   [x] All visualizations display properly
-   [x] Headers and formatting render correctly
-   [x] File is named successfully posted to your website

------------------------------------------------------------------------
